01_Vegetation_cover

### 2022-10-07 01_Species_abundance.Rmd
add NWSS data

Use Sample raster values to compute landcover types for vegetation
plots. Export as Veg_type_zoneB_sites.csv and Veg_type_zoneA_sites.csv

## Prepare data

@@

```{r load and prepare vegetation data}
# Load in vegetation type data for the plots
zone_b_site <- read.csv("Data/20220106-Veg_type_zoneB_sites.csv")

# Make tidy
zone_b_site <- zone_b_site %>% 
  # create seperate site and plot column
  separate(name, c("sitename", "plot")) %>% 
  # recode veg_type
  mutate(veg_type = as.factor(rvalue_1)) %>% 
  dplyr::select(sitename, plot, lat, lon, veg_type) %>% 
  mutate(veg_type = recode(veg_type, 
                           "1" = "C1: Water",
                           "2" = "D1: Raised and blanket bogs",
                           "3" = "D2: Valley mires, poor fens and transition mires",
                           "4" = "D4: Base-rich fens and calcareous spring mires", 
                           "5" = "E1: Dry grasslands",
                           "6" = "E2: Mesic grasslands",
                           "7" = "E3: Seasonally wet and wet grasslands",
                           "8" = "E4: Alpine and subalpine grasslands",
                           "9" = "E5: Woodland fringes and clearings and tall forb stands",
                           "10"= "F2: Arctic, alpine and subalpine scrub",
                           "11" = "F3: Temperate and mediterranean-montane scrub",
                           "12"= "F4: Temperate shrub heathland",
                           "13" = "F9: Riverine and fen scrubs",
                           "14"= "G1: Broadleaved deciduous woodland",
                           "15" = "G3: Coniferous woodland",
                           "16" = "G4: Mixed deciduous and coniferous woodland",
                           "17" = "G5: Line of trees, amall anthropogenic woodlands, recently felled woodland, early-stage woodland and coppice",
                           "18" = "H2: Screes",
                           "19" = "Inland cliffs, rock pavements and outcrops",
                           "20" = "I1: Arable land and market gardens",
                           "21" = "J: Builtup (and Bare before post)",
                           "23" = "O: Bare (after post)"))

# load in vegetation data
spec_b_raw <- read_xlsx("Data/species_zone_B.xlsx")

# make tidy
spec_b <- spec_b_raw %>% 
  pivot_longer(contains("plot"), names_to = "plot", values_to = "abun",
               values_drop_na = TRUE) %>% 
  mutate(plot = str_remove(plot, "plot_ ")) %>% 
  dplyr::select(sitename = Site, plot, species = Species, abun) %>% 
  # join with vegetation type data
  left_join(zone_b_site, by = c("sitename", "plot"))
```

```{r calculate vegetation cover}
veg_cover_bc <- spec_b %>% 
  # remove moss
  filter(!species %in% c("Moss", "Sphagnum", "Polytrichum commune",
                         "Sphagnum squarrosum")) %>% 
  # calculate avarage species cover per vegetation type using the plots belonging to that vegetation type
  group_by(sitename, veg_type, species) %>% 
  summarise(abun = mean(abun)) %>% 
  # join with vegetation cover data
  full_join(veg_type, by = c("sitename", "veg_type")) %>% 
  # weight species abundance by veg cover
  mutate(spec_abun_b = abun*veg_cov_b, spec_abun_c = abun*veg_cov_c) %>% 
  # drop unneeded columns
  dplyr::select(sitename, veg_type, species, veg_cov_b, veg_cov_c,
                spec_abun_b,spec_abun_c)

# print table with missing species data for vegetation types
missing_vegtype <- veg_cover_bc %>% 
  filter(is.na(species)) %>% 
  group_by(sitename, veg_type) %>% 
  summarise(cov_b = round(sum(veg_cov_b), 2), cov_c = round(sum(veg_cov_c), 2)) %>%
  filter(!cov_b == 0|!cov_c == 0)

# for missing data, use species data from other plots
missing <- spec_b %>% 
  # remove moss
  filter(!species %in% c("Moss", "Sphagnum", "Polytrichum commune",
                         "Sphagnum squarrosum")) %>% 
  filter(veg_type %in% missing_vegtype$veg_type) %>% 
  group_by(veg_type, species) %>% 
  summarise(abun = mean(abun))

veg_cover_bc_na <- veg_cover_bc %>% 
  filter(is.na(species)) %>%
  dplyr::select(-species) %>% 
  left_join(missing, by = "veg_type") %>% 
  mutate(spec_abun_b = abun*veg_cov_b, spec_abun_c = abun*veg_cov_c) %>% 
  dplyr::select(sitename, veg_type, species, veg_cov_b, veg_cov_c, spec_abun_b,
         spec_abun_c)

veg_cover_bc <-  veg_cover_bc %>% 
  filter(!is.na(species)) %>% 
  bind_rows(., veg_cover_bc_na) %>% 
  arrange(sitename, veg_type, species)
```

@@

```{r standardize species names}
# load in generalized species list, spelling checked and harmonized with
# LCVP
harm <- read_xlsx("Data/Harmonization_table_species_names.xlsx") 
harm <- harm %>% 
  mutate(stand.spec = str_remove(stand.spec, pattern = " sp\\."))
spec_a <- spec_a %>% 
  left_join(harm, by = "species")
veg_cover_bc <- veg_cover_bc %>% 
  left_join(harm, by = "species")

@@

# print table with missing species data for vegetation types
missing_vegtype <- spec_cover %>% 
  filter(is.na(stand.spec)) %>% 
  group_by(sitename, LABEL_DE) %>% 
  summarise(cov_b = round(sum(veg_cov_b), 2), 
            cov_c = round(sum(veg_cov_c), 2)) %>%
  filter(!cov_b == 0|!cov_c == 0)

>>>>

Use Sample raster values to compute landcover types for vegetation
plots. Export as Veg_type_zoneB_sites.csv and Veg_type_zoneA_sites.csv,

Canopy data for zoneB was extracted for the zone B sites from the
Scottish Native Woodland survey:
<https://forestry.gov.scot/forests-environment/biodiversity/native-woodlands/native-woodland-survey-of-scotland-nwss>

Patterson, G., Nelson, D., Robertson, P. and Tullis, J., 2014.
*Scotland's Native Woodlands: Results from the Native Woodland Survey of
Scotland*. [online] Edinburgh: Forestry Commission Scotland. Available
at:
<https://forestry.gov.scot/publications/74-scotland-s-native-woodlands-results-from-the-native-woodland-survey-of-scotland>
[Accessed 3 October 2022].

-   Join NWSS woodland cover shape file with Species Structures table -
    by SCPTDATA_I

-   Union zone buffer (B or C separately) and NWSS woodland vector layer

-   remove all that don't overlap with the zone buffer

-   calculate area of each polygon with field calculator \$area

-   export to csv

@@

```{r load and prepare species data}
# Load in vegetation type data for the plots
zone_b_site <- read.csv("Data/20220106-Veg_type_zoneB_sites.csv")
spec_b_raw <- read_xlsx("Data/species_zone_B.xlsx") # load in species data

# Make vegetation type tidy
zone_b_site <- zone_b_site %>% 
  # create seperate site and plot column
  separate(name, c("sitename", "plot")) %>% 
  # recode veg_type
  mutate(veg_type = as.factor(rvalue_1)) %>% 
  dplyr::select(sitename, plot, lat, lon, veg_type) %>% 
  mutate(veg_type = recode(veg_type, 
                           "1" = "C1: Water",
                           "2" = "D1: Raised and blanket bogs",
                           "3" = "D2: Valley mires, poor fens and transition mires",
                           "4" = "D4: Base-rich fens and calcareous spring mires", 
                           "5" = "E1: Dry grasslands",
                           "6" = "E2: Mesic grasslands",
                           "7" = "E3: Seasonally wet and wet grasslands",
                           "8" = "E4: Alpine and subalpine grasslands",
                           "9" = "E5: Woodland fringes and clearings and tall forb stands",
                           "10" = "F2: Arctic, alpine and subalpine scrub",
                           "11"  = "F3: Temperate and mediterranean-montane scrub",
                           "12" = "F4: Temperate shrub heathland",
                           "13" = "F9: Riverine and fen scrubs",
                           "14" = "G1: Broadleaved deciduous woodland",
                           "15" = "G3: Coniferous woodland",
                           "16" = "G4: Mixed deciduous and coniferous woodland",
                           "17" = "G5: Line of trees, amall anthropogenic woodlands, recently felled woodland, early-stage woodland and coppice",
                           "18" = "H2: Screes",
                           "19" = "Inland cliffs, rock pavements and outcrops",
                           "20" = "I1: Arable land and market gardens",
                           "21" = "J: Builtup (and Bare before post)",
                           "23" = "O: Bare (after post)"))

# Tidy and join 
spec_b <- spec_b_raw %>% 
  pivot_longer(contains("plot"), names_to = "plot", values_to = "abun",
               values_drop_na = TRUE) %>% 
  mutate(plot = str_remove(plot, "plot_ ")) %>% 
  dplyr::select(sitename = Site, plot, species = Species, abun) %>% 
  # join with vegetation type data
  left_join(zone_b_site, by = c("sitename", "plot")) %>% 
  # filter trees from species survey (will be covered by NWSS)
  filter(!species %in% c("Betula pendula", "Pinus sylvestris", 
                        "Juniperus communis", "Salix sp.",
                        "Larix decidua", "Salix repens"))
```

tree abundance

```{r}
canop_b_raw <- 
  read.csv("Data/NWSS_zoneB_ID_forest_cover.csv") %>% 
  # clean column names
  rename_with(~str_remove(., "NWSS_Species_Structures_5718542285393555851.â...NWSS_Species_Structures_"),
              matches("NWSS_Species_Structures_5718542285393555851.â...NWSS_Species_Structures_"))
canop_c_raw <- 
  read.csv("Data/NWSS_zoneC_ID_forest_cover.csv") %>% 
  # clean column names
  rename_with(~str_remove(., "NWSS_Species_Structures_5718542285393555851.â...NWSS_Species_Structures_"),
              matches("NWSS_Species_Structures_5718542285393555851.â...NWSS_Species_Structures_"))
NWSS_sp <- 
  read.csv("Data/NWSS_Species_Structures.csv") 
NWSS <- 
  read.csv("Data/NWSS_basis.csv") %>%  # inventory of NWSS 
  # clean column names
  rename_with(~str_remove(., "NWSS_Species_Structures_"),
              matches("NWSS_Species_Structures_"))

# calculate average species abundance of species per habitat type
# species percentage of NWSS (not available for all polygon)
habitats <- canop_c_raw$DOM_HABITA %>% unique
sp_habitat <- NWSS %>% 
  filter(DOM_HABITA %in% habitats) %>% 
  group_by(DOM_HABITA, SPECIES) %>% 
  summarise(mean_pct = mean(SP_STR_PCT))

# - calculate woodland cover for every site
# - weight woodland cover by mean abundance of species per type
canop_b <- canop_b_raw %>% 
  # select relevant columns
  dplyr::select(sitename = name, SCPTDATA_I, TYPE, DOM_HABITA, woodland_area = Woodland_area) %>% 
  # fill in area without woodland
  mutate(TYPE = if_else(TYPE == "", "Open land habitat", TYPE),
         # rename site names
         sitename = str_remove(sitename, "-A")) %>% 
  # calculate woodland type percentage
  group_by(sitename) %>% 
  # calculate percentages
  mutate(total_area = sum(woodland_area)) %>% 
  mutate(woodland_percentage = woodland_area/total_area) %>% 
  # join with species abundance
  left_join(sp_habitat, by = "DOM_HABITA") %>% 
  # create column with scientific species name
  mutate(species = recode(SPECIES, "Common alder*" = "Alnus glutinosa",
                          "Rowan *" = "Sorbus aucuparia",
                          "Sitka spruce" = "Picea sitchensis",
                          "European larch" = "Larix decidua",
                          "Lodgepole pine" = "Pinus contorta",
                          "Downy birch *" = "Betula pubescens",
                          "Scots pine *" = "Pinus sylvestris",
                          "Hazel *" = "Corylus avellana",
                          "Eared willow *+" = "Salix aurita", 
                          "Grey willow *" = "Salix cinerea", 
                          "Hybrid larch" = "Larix decidua", 
                          "Shrubs - preferentially browsed +" = "Unidentified",
                          "Dwarf juniper *+" = "Juniperus communis", 
                          "Silver birch *" = "Betula pendula", 
                          "Beech" = "Fagus sylvatica", 
                          "Lawsons cypress" = "Chamaecyparis lawsoniana", 
                          "Common alder*" = "Alnus glutinosa", 
                          "Douglas fir" = "Pseudotsuga menziesii", 
                          "Yew *" = "Taxus baccata", 
                          "Upright juniper * +" = "Juniperus communis", 
                          "Norway spruce" = "Picea abies", 
                          "Bird cherry *" = "Prunus padus", 
                          "Intermediate oak *" = "Quercus sp.", 
                          "Sessile oak *" = "Quercus petraea", 
                          "Ash *" = "Fraxinus excelsior", 
                          "Holly *" = "Ilex aquifolium", 
                          "Aspen *" = "Populus tremula", 
                          "Goat willow*" = "Salix caprea", 
                          "Elder *" = "Sambucus nigra", 
                          "Broom *+" = "Cytisus scoparius", 
                          "Other non-native trees/shrubs" = "Unidentified", 
                          "Sycamore" = "Acer pseudoplatanus", 
                          "Grand fir" = "Abies grandis", 
                          "Gorse *+" = "Ulex europaeus", 
                          "Gean wild cherry *" = "Prunus avium",
                          "Austrian pine" = "Pinus nigra",
                          "Blackthorn *+" = "Prunus spinosa",
                          "Birch *" = "Betula sp.",
                          "Coast redwood" = "Sequoia sempervirens",
                          "Bramble *+" = "Rubus fruticosus",
                          "Crab apple" = "Malus sylvestris",
                          "Dog rose *+" = "Rosa canina",
                          "Geulder rose *+" = "Viburnum opulus",
                          "Grey alder" = "Alnus incana",
                          "Hawthorn *" = "Crataegus monogyna",
                          "Field maple" = "Acer campestris",
                          "Hornbeam" = "Carpinus betulus",               
                          "Horse chestnut" = "Aesculus hippocastanum",
                          "Japanese knotweed +" = "Reynoutria japonica",
                          "Japanese larch" = "Larix kaempferi",
                          "Leyland cypress" = "Cupressus sp.",        
                          "Lime all" = "Tilia sp.",
                          "Noble fir" = "Abies procera",
                          "Non-native willow" = "Salix sp.",
                          "Norway maple" = "Acer platanoides",       
                          "Nothofagus all" = "Nothofagus sp.",
                          "Omorika spruce" = "Picea omorika",         
                          "Other native willows *" = "Salix sp.",
                          "Other pines" = "Pinus sp.",
                          "Other whitebeam"  = "Sorbus aria",       
                          "Pedunculate oak *" = "Quercus robur",     
                          "Poplar" = "Populus sp.",
                          "Red oak" = "Quercus rubra",               
                          "Rhododendron ponticum +" = "Rhododendron ponticum",
                          "Scots pine" = "Pinus sylvestris",
                          "Silver fir" = "Abies alba",
                          "Snowberry+" = "Symphoricarpos sp.",
                          "Sweet chestnut" = "Castanea sativa",
                          "Western hemlock" = "Tsuga heterophylla",
                          "Western red cedar" = "Thuja plicata",
                          "Wild service tree" = "Sorbus torminalis",      
                          "Wych elm *" = "Ulmus glabra",
                          "English elm" = "Ulmus procera",
                          "Japanese cedar" = "Cryptomeria japonica",
                          "Montane willow all *+" = "Salix sp."
                          )) %>% 
  # filter non trees
  mutate(growthform = case_when(str_detect(SPECIES, "\\+") ~ "shrub",
                                SPECIES %in% c("Japanese knotweed +") ~ "herb",
                                TRUE ~ "tree")) %>% 
  filter(growthform == "tree",
         !species %in% c("Unidentified","")) %>% 
  # weight species abundance by woodland percentage
  group_by(sitename, species) %>% 
  mutate(sp_abun = mean_pct * woodland_percentage) %>% 
  summarise(spec_abun_b = sum(sp_abun))

# PAWS =  Planted Woods on Ancient Woodland Sites
canop_c <- canop_c_raw %>% 
  # select relevant columns
  dplyr::select(sitename = name, SCPTDATA_I, TYPE, DOM_HABITA, woodland_area = Woodland_area) %>% 
  # fill in area without woodland
  mutate(TYPE = if_else(TYPE == "", "Open land habitat", TYPE),
         # rename site names
         sitename = str_remove(sitename, "-A")) %>% 
  # calculate woodland type percentage
  group_by(sitename) %>% 
  # calculate percentages
  mutate(total_area = sum(woodland_area)) %>% 
  mutate(woodland_percentage = woodland_area/total_area) %>% 
  # join with species abundance
  left_join(sp_habitat, by = "DOM_HABITA") %>% 
  # create column with scientific species name
  mutate(species = recode(SPECIES, "Common alder*" = "Alnus glutinosa",
                          "Rowan *" = "Sorbus aucuparia",
                          "Sitka spruce" = "Picea sitchensis",
                          "European larch" = "Larix decidua",
                          "Lodgepole pine" = "Pinus contorta",
                          "Downy birch *" = "Betula pubescens",
                          "Scots pine *" = "Pinus sylvestris",
                          "Hazel *" = "Corylus avellana",
                          "Eared willow *+" = "Salix aurita", 
                          "Grey willow *" = "Salix cinerea", 
                          "Hybrid larch" = "Larix decidua", 
                          "Shrubs - preferentially browsed +" = "Unidentified",
                          "Dwarf juniper *+" = "Juniperus communis", 
                          "Silver birch *" = "Betula pendula", 
                          "Beech" = "Fagus sylvatica", 
                          "Lawsons cypress" = "Chamaecyparis lawsoniana", 
                          "Common alder*" = "Alnus glutinosa", 
                          "Douglas fir" = "Pseudotsuga menziesii", 
                          "Yew *" = "Taxus baccata", 
                          "Upright juniper * +" = "Juniperus communis", 
                          "Norway spruce" = "Picea abies", 
                          "Bird cherry *" = "Prunus padus", 
                          "Intermediate oak *" = "Quercus sp.", 
                          "Sessile oak *" = "Quercus petraea", 
                          "Ash *" = "Fraxinus excelsior", 
                          "Holly *" = "Ilex aquifolium", 
                          "Aspen *" = "Populus tremula", 
                          "Goat willow*" = "Salix caprea", 
                          "Elder *" = "Sambucus nigra", 
                          "Broom *+" = "Cytisus scoparius", 
                          "Other non-native trees/shrubs" = "Unidentified", 
                          "Sycamore" = "Acer pseudoplatanus", 
                          "Grand fir" = "Abies grandis", 
                          "Gorse *+" = "Ulex europaeus", 
                          "Gean wild cherry *" = "Prunus avium",
                          "Austrian pine" = "Pinus nigra",
                          "Blackthorn *+" = "Prunus spinosa",
                          "Birch *" = "Betula sp.",
                          "Coast redwood" = "Sequoia sempervirens",
                          "Bramble *+" = "Rubus fruticosus",
                          "Crab apple" = "Malus sylvestris",
                          "Dog rose *+" = "Rosa canina",
                          "Geulder rose *+" = "Viburnum opulus",
                          "Grey alder" = "Alnus incana",
                          "Hawthorn *" = "Crataegus monogyna",
                          "Field maple" = "Acer campestris",
                          "Hornbeam" = "Carpinus betulsu",               
                          "Horse chestnut" = "Aesculus hippocastanum",
                          "Japanese knotweed +" = "Reynoutria japonica",
                          "Japanese larch" = "Larix kaempferi",
                          "Leyland cypress" = "Cupressus sp.",        
                          "Lime all" = "Tilia sp.",
                          "Noble fir" = "Abies procera",
                          "Non-native willow" = "Salix sp.",
                          "Norway maple" = "Acer platanoides",       
                          "Nothofagus all" = "Nothofagus sp.",
                          "Omorika spruce" = "Picea omorika",         
                          "Other native willows *" = "Salix sp.",
                          "Other pines" = "Pinus sp.",
                          "Other whitebeam"  = "Sorbus aria",       
                          "Pedunculate oak *" = "Quercus robur",     
                          "Poplar" = "Populus sp.",
                          "Red oak" = "Quercus rubra",               
                          "Rhododendron ponticum +" = "Rhododendron ponticum",
                          "Scots pine" = "Pinus sylvestris",
                          "Silver fir" = "Abies alba",
                          "Snowberry+" = "Symphoricarpos sp.",
                          "Sweet chestnut" = "Castanea sativa",
                          "Western hemlock" = "Tsuga heterophylla",
                          "Western red cedar" = "Thuja plicata",
                          "Wild service tree" = "Sorbus torminalis",      
                          "Wych elm *" = "Ulmus glabra",
                          "English elm" = "Ulmus procera",
                          "Japanese cedar" = "Cryptomeria japonica",
                          "Montane willow all *+" = "Salix sp.",
                          "Arran whitebeam *" = "Sorbus arranensis"
         )) %>% 
  # filter non trees
  mutate(growthform = case_when(str_detect(SPECIES, "\\+") ~ "shrub",
                                SPECIES %in% c("Japanese knotweed +") ~ "herb",
                                TRUE ~ "tree")) %>% 
  filter(growthform == "tree",
       !species %in% c("Unidentified","")) %>% 
  # weight species abundance by woodland percentage
  group_by(sitename, species) %>% 
  mutate(sp_abun = mean_pct * woodland_percentage) %>% 
  summarise(spec_abun_c = sum(sp_abun))

# join df's from both zones
canop <- canop_b %>% 
  left_join(canop_c, by = c("sitename","species"))
```

```{r calculate vegetation cover}
veg_cover_bc <- spec_b %>% 
  # remove moss
  filter(!species %in% c("Moss", "Sphagnum", "Polytrichum commune",
                         "Sphagnum squarrosum")) %>% 
  # calculate average species cover per vegetation type using the plots belonging to that vegetation type
  group_by(sitename, veg_type, species) %>% 
  summarise(abun = mean(abun)) %>% 
  # join with vegetation cover data
  full_join(veg_type, by = c("sitename", "veg_type"))

# print table with missing species data for vegetation types
missing_vegtype <- veg_cover_bc %>% 
  filter(is.na(species)) %>% 
  group_by(sitename, veg_type) %>% 
  summarise(cov_b = round(sum(veg_cov_b), 2), cov_c = round(sum(veg_cov_c), 2)) %>%
  filter(!cov_b == 0 | !cov_c == 0)

# for missing data, use species data from other plots
missing <- spec_b %>% 
  # remove moss
  filter(!species %in% c("Moss", "Sphagnum", "Polytrichum commune",
                         "Sphagnum squarrosum")) %>% 
  filter(veg_type %in% missing_vegtype$veg_type) %>% 
  group_by(veg_type, species) %>% 
  summarise(abun = mean(abun))
veg_cover_bc_na <- veg_cover_bc %>% 
  filter(is.na(species)) %>%
  dplyr::select(-species, -abun) %>% 
  left_join(missing, by = "veg_type")

veg_cover_bc <-  veg_cover_bc %>% 
  filter(!is.na(species)) %>% 
  # bind with df with missing veg types
  bind_rows(., veg_cover_bc_na) %>% 
  # weight species abundance by veg cover
  group_by(sitename, species) %>% 
  mutate(spec_abun_b = abun*veg_cov_b, spec_abun_c = abun*veg_cov_c) %>%
  summarise(spec_abun_b = sum(spec_abun_b),
            spec_abun_c = sum(spec_abun_b)) %>% 
  # bind with tree species data
  bind_rows(canop) %>% 
  arrange(sitename, species)
```

@@

```{r standardize species names}
# load in generalized species list, spelling checked and harmonized with LCVP
harm <- read_xlsx("Data/Harmonization_table_species_names.xlsx") 
harm <- harm %>% 
  mutate(stand.spec = str_remove(stand.spec, pattern = " sp\\."))

spec_a <- spec_a %>% 
  left_join(harm, by = "species")
veg_cover_bc <- veg_cover_bc %>% 
  left_join(harm, by = "species") %>% 
  filter(!is.na(stand.spec))

@@

# print table with missing species data for vegetation types
missing_vegtype <- spec_cover %>% 
  filter(is.na(stand.spec)) %>% 
  group_by(sitename, LABEL_DE) %>% 
  summarise(cov_b = round(sum(veg_cov_b), 2), 
            cov_c = round(sum(veg_cov_c), 2)) %>%
  filter(!cov_b == 0 |!cov_c == 0)
